---
title: "Random Forest Weather Data"
author: "Samantha Bothwell"
date: "6/11/2019"
output: pdf_document
---
SUMMARY OF RANDOM FOREST

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load packages}
# Load important packages
library(tidyr)
library(randomforest)
```

```{r clean data}
# First we read in the data and clean it
setwd("/Users/sbothwell/Desktop/DATAS/My_material")
dat = read.csv("weatherdata.csv")

# Rename columns
colnames(dat) = c("DateTime(MST)","Temp(degF)","Humidity%","DewPt(degF)","Wind(mph)",
                  "WindDir(degNorth)","Gust(mph)","GustDir(degNorth)","Pressure(Hg)",
                  "Solar(W/m^2)","Percipitation")

# Remove second row 
dat = dat[-1,]

# Split date and time
dat = dat %>% separate(`DateTime(MST)`, c("Date","Time(MST)"), " ")

# Split date
dat = dat %>% separate(Date, c("Year","Month","Day"), "-")

# Change numeric month to character
dat$Month <- as.numeric(dat$Month)
dat$Month <- month.abb[dat$Month]
```

Now that our data is clean, let's walk through using the Random Forest Algorithm.
First we will make data sets for testing and training. These are random samples from our original data. We will use the training data to make the model and the testing data to evaluate how effective our model is at making predictions. For this data, I will make my training data 75% of the original data set and testing data the other 25% of the original data set. 
```{r make data sets}
set.seed(2019) # set seed for reproducible results

## 75% of the sample size
smp_size <- floor(0.75 * nrow(dat))
train_ind <- sample(seq_len(nrow(dat)), size = smp_size)

train <- dat[train_ind, ]
test <- dat[-train_ind, ]
```



